Nextstrain: real-time tracking of pathogen evolution

### 2018-04-16 replyToReviewer.tex

\documentclass[12pt,a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\begin{document}

Please find our reply to the points raised by reviewer 1.

\begin{quotation}
General comments:
In this well-written and timely Applications Note, Hadfield and colleagues present the Nextstrain platform – a web-based visualization and analysis platform for viral genome phylogenies. This reviewer has used the site on many occasions to explore viral genomic data, and can attest to the fact that Nextstrain is an incredibly useful resource to the viral genomics and public health communities. Its thoughtful, intuitive interface makes the tool easy to use, even for individuals with no formal bioinformatics training. Nextstrain has played a role in a number of recent outbreaks of public health importance, from Ebola to Zika, and is an excellent resource for annual surveillance of influenza viruses. The platform's commitment to open science is exceptional, and demonstrates to the viral genomics community how powerful data-sharing can be.

\paragraph{}
This is an excellent Applications Note describing a fantastic tool, and I have no hesitations in recommending the manuscript for publication. I have no major revisions to suggest, and only very minor style suggestions that the authors may take or leave:

\end{quotation}

1. Page 1, Line 39: "ongoing and ever-present" is somewhat redundant; I'd just go with "ever-present"
2. Page 1, Line 40: perhaps change "recent epidemics" to "recent events" so "epidemics" isn't repeated 3x in the sentence.

3. Page 1, Line 42: it might be helpful to point the reader to a general reference on phylodynamics, assuming the reference limit hasn't already been reached.

4. Page 1, Line 43 (column 2): could references to the three databases (papers or URLs) be included here?

5. Page 2, Line 35: if it can be explained in a very short sentence, it might be interesting to mention the caveat of "nearly all outbreaks" - which ones aren't amenable to Nextstrain-ification?

6. Page 2, Line 51: it should be noted that the transmission inference depends on how well the outbreak in question was sampled - if data is missing, important transmission links could be missed.

7. Page 2, Line 20 (column 20): is there an example of one of these mutations that could be mentioned here?

There is a fair amount of passive voice in the manuscript - popping the text into readable.io to find these instances and changing the construction to make them active will make the writing a little more lively (and it usually frees up some words in the word count).

\end{document}

### 2018-04-16 replyToReviewer.tex

\begin{quotation}
General comments:
In this well-written and timely Applications Note, Hadfield and colleagues present the Nextstrain platform – a web-based visualization and analysis platform for viral genome phylogenies. This reviewer has used the site on many occasions to explore viral genomic data, and can attest to the fact that Nextstrain is an incredibly useful resource to the viral genomics and public health communities. Its thoughtful, intuitive interface makes the tool easy to use, even for individuals with no formal bioinformatics training. Nextstrain has played a role in a number of recent outbreaks of public health importance, from Ebola to Zika, and is an excellent resource for annual surveillance of influenza viruses. The platform's commitment to open science is exceptional, and demonstrates to the viral genomics community how powerful data-sharing can be.

\paragraph{}
This is an excellent Applications Note describing a fantastic tool, and I have no hesitations in recommending the manuscript for publication. I have no major revisions to suggest, and only very minor style suggestions that the authors may take or leave:

\end{quotation}

We thank the reviewer for these kind words!

\begin{quote}
1. Page 1, Line 39: "ongoing and ever-present" is somewhat redundant; I'd just go with "ever-present"

2. Page 1, Line 40: perhaps change "recent epidemics" to "recent events" so "epidemics" isn't repeated 3x in the sentence.
\end{quote}

We have changed the wording of the manuscript to address these points.

\begin{quote}
3. Page 1, Line 42: it might be helpful to point the reader to a general reference on phylodynamics, assuming the reference limit hasn't already been reached.
\end{quote}

### 2018-04-16 whitepaper.tex
changes for submission
points 1 & 2

\abstract{
\textbf{Summary:}
Understanding the spread and evolution of pathogens is important for effective public health measures and surveillance.
%Genomic epidemiology, aided by the often high rates of viral evolution and ability to rapidly disseminate around the world, improves our understanding of such diseases.
Nextstrain consists of a database of viral genomes, a bioinformatics pipeline for phylodynamics analysis, and an interactive visualisation platform.
Together these present a real-time view into the evolution and spread of a range of viral pathogens of high public health importance.
The visualization integrates sequence data with other data types such as geographic information, serology, or host species.
Nextstrain compiles our current understanding into a single accessible location, publicly available for use by health professionals, epidemiologists, virologists and the public alike.
\\
\textbf{Availability and implementation:}
All code (predominantly JavaScript and Python) is freely available from \href{https://github.com/nextstrain}{github.com/nextstrain} and the web-application is available at \href{http://nextstrain.org}{nextstrain.org}.
\\
\textbf{Contact:}
\href{jhadfiel@fredhutch.org}{jhadfiel@fredhutch.org}, \href{tbedford@fredhutch.org}{tbedford@fredhutch.org}, \href{richard.neher@unibas.ch}{richard.neher@unibas.ch}
\\
}

@@

Viral pathogens pose an ongoing and ever-present danger to global human health, with recent epidemics such as the West-African Ebola epidemic and the ongoing Zika epidemic in the Americas highlighting the risk.
The rapid evolution of these viruses allows inference of epidemic history from genomic data.
Such analyses are often done in isolation, and may lack the spatial or temporal context in which to best interpret the results, which impedes understanding \citep{pybus2013evolutionary}.
Furthermore, the results of analyses are often not made available to the public or health bodies until after publication, which may be too late to effect change in policy or understanding.
We have developed Nextstrain to visualise outbreaks in as close to real time as possible.
Whilst currently encompassing a selection of viruses, extension to non-viral pathogens is forthcoming.
% This platform consists of three components: the collection and updating of available sequences from a range of public and private sources, rapid and automated analysis pipelines and an interactive web-application by which to interrogate the results.

@@

\begin{figure}[!tpb]
\centerline{\includegraphics[width=0.48\textwidth]{figures/nextstrain}}
\caption{Genomic epidemiology of Zika virus as of Oct 2017 (live display at \href{http://nextstrain.org/zika}{nextstrain.org/zika}). The main interface consists of three linked panels --- a phylogenetic tree, geographic transmissions, and the genetic diversity across the genome.}
\label{nextstrain}
\end{figure}

@@

% please add acknowledgment if applicable.
%\section*{Acknowledgements}
%We would like to thank...
%
%\vspace*{-12pt}

>>>>

\abstract{
\textbf{Summary:}
Understanding the spread and evolution of pathogens is important for effective public health measures and surveillance.
Nextstrain consists of a database of viral genomes, a bioinformatics pipeline for phylodynamics analysis, and an interactive visualisation platform.
Together these present a real-time view into the evolution and spread of a range of viral pathogens of high public health importance.
The visualization integrates sequence data with other data types such as geographic information, serology, or host species.
Nextstrain compiles our current understanding into a single accessible location, publicly available for use by health professionals, epidemiologists, virologists and the public alike.
\\
\textbf{Availability and implementation:}
All code (predominantly JavaScript and Python) is freely available from \href{https://github.com/nextstrain}{github.com/nextstrain} and the web-application is available at \href{http://nextstrain.org}{nextstrain.org}.
\\
\textbf{Contact:}
\href{jhadfiel@fredhutch.org}{jhadfiel@fredhutch.org}, \href{tbedford@fredhutch.org}{tbedford@fredhutch.org}, \href{richard.neher@unibas.ch}{richard.neher@unibas.ch}
\\
}

@@

Viral pathogens pose an ever-present danger to global human health, with recent events such as the West-African Ebola epidemic and the ongoing Zika epidemic in the Americas highlighting the risk.
The rapid evolution of these viruses allows inference of epidemic history from genomic data.
Such analyses are often done in isolation, and may lack the spatial or temporal context in which to best interpret the results, which impedes understanding \citep{pybus2013evolutionary}.
Furthermore, the results of analyses are often not made available to the public or health bodies until after publication, which may be too late to effect change in policy or understanding.
We have developed Nextstrain to visualise outbreaks in as close to real time as possible.
Whilst currently encompassing a selection of viruses, extension to non-viral pathogens is forthcoming.

@@

\begin{figure}[!tpb]
\centerline{\includegraphics[width=0.48\textwidth]{figure}}
\caption{Genomic epidemiology of Zika virus as of Oct 2017 (live display at \href{http://nextstrain.org/zika}{nextstrain.org/zika}). The main interface consists of three linked panels --- a phylogenetic tree, geographic transmissions, and the genetic diversity across the genome.}
\label{nextstrain}
\end{figure}

@@

% please add acknowledgment if applicable.
%\section*{Acknowledgements}
%\vspace*{-12pt}
